---
title: "Data607 - Project 2: Data Transformation with Astronomical Data"
author: "Anthony Josue Roman"
date: "2024-10-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include=FALSE}
library(readr)
library(RCurl)
library(stringr)
library(dplyr)
library(tidyr)
library(tidyverse)
library(ggplot2)
library(knitr)
library(kableExtra)
```

This assignment will use the following packages:

```{r eval=FALSE}
library(readr)
library(RCurl)
library(stringr)
library(dplyr)
library(tidyr)
library(tidyverse)
library(ggplot2)
library(knitr)
library(kableExtra)
```

## Introduction

The following project will utilize three datasets I have obtained from either simulation or from the Sloan Digital Sky survey data (fun fact my mentor worked on this sky survey data - Donald P Schneider). The first dataset will be exoplanet dataset from [Sloan Digital Sky Survey](https://exoplanetarchive.ipac.caltech.edu/cgi-bin/TblView/nph-tblView?app=ExoTbls&config=PS), the second dataset will be from [Gaia Archeive](https://gea.esac.esa.int/archive/) for Stellar Evolution, and the last one will be from [EZ-Webb](http://user.astro.wisc.edu/~townsend/static.php?ref=ez-web#toc-Submit_a_Calculation) where I used **three** different calculations from 3 different stellar masses with the same Metallicity.

## Initializing datasets

